---
title: "Data612 - Project 3"
author: "Anna Moy & Natalie Kalukeerthie"
date: "2025-06-22"
output: html_document
---

# What is Singular Value Decomposition (SVD)?

A method used in linear algebra to decompose a matrix into 3 simpler matrices which makes it easier to analyze and manipulate the data. 


he Singular Value Decomposition (SVD) of a matrix \( R \) is:

\[
R \approx U \Sigma V^T
\]

Where:

- \( R \) is the user–item rating matrix,
- \( U \) is the matrix of user latent features,
- \( \Sigma \) (or \( D \)) is the diagonal matrix of singular values,
- \( V^T \) is the transpose of the item latent feature matrix.

```{r, warning=FALSE, message=FALSE}
# Load libraries
library(recommenderlab)
library(irlba)
```

```{r}
# Load MovieLens data
data("MovieLense")
ratings <- MovieLense
```

```{r}
# Create a smaller subset
set.seed(42)
small_ratings <- ratings[1:100, 1:100]  # 100 users, 100 items

# Convert to matrix for SVD
rating_mat <- as(small_ratings, "matrix")

# Step 1: Replace NAs with row means (with fallback to global mean)
row_means <- rowMeans(rating_mat, na.rm = TRUE)
global_mean <- mean(rating_mat, na.rm = TRUE)
row_means[is.na(row_means)] <- global_mean  # handle rows with all NA

rating_imputed <- rating_mat
for (i in 1:nrow(rating_imputed)) {
  rating_imputed[i, is.na(rating_imputed[i, ])] <- row_means[i]
}

# Step 2: Proper row-centering using sweep()
rating_centered <- sweep(rating_imputed, 1, row_means)

# Step 3: Apply truncated SVD
svd_result <- irlba(rating_centered, nv = 20)

# Step 4: Reconstruct approximate ratings
approx_ratings <- svd_result$u %*% diag(svd_result$d) %*% t(svd_result$v)

# Step 5: Add back row means using sweep
predicted_ratings <- sweep(approx_ratings, 1, row_means, FUN = "+")

# Step 6: View predictions
head(predicted_ratings[, 1:5])

```